Towards Artificial Intelligence for Metaphysics: \\ the Case of the Inconsistency in G\"odel's Ontological Argument

### 2016-01-08 Papers/InProgress/IJCAI/paper.tex
Move figure
Improve introduction

\begin{abstract}
  This paper discusses the discovery of the inconsistency in G\"odel's ontological argument as a success story for artificial intelligence. Despite the popularity of the argument since the appearance of G\"odel's manuscript in 1970, the inconsistency of the axioms used in the argument remained unnoticed until 2013, when it was detected automatically by the higher-order theorem prover LEO-II. Understanding and verifying the refutation generated by the prover turned out to be a time-consuming task, and its completion, as reported here, required the reconstruction of the refutation in the Isabelle proof assistant. The development of an improved syntactical hiding for the embedding technique, utilizing Isabelle's binding notation mechanism, allows the refutation to be presented in a human-friendly way, suitable for non-experts in the technicalities of the higher-order theorem proving. This brings us a step closer to wider adoption of logic-based artificial intelligence tools by philosophers.
\end{abstract}

@@

The novel contributions reported in this paper include the following:

\begin{itemize}
\item A detailed analysis of the inconsistency of G\"{o}del's
  \shortcite{GoedelNotes} original version of the proof is presented
  for higher-order modal logics KB and K (with possibilist and
  actualist first-order quantifiers). The extraction, reconstruction
  and verification of an informal, human intuitive argument has been
  an open problem since the first detection of this inconsistency by
  Benzm\"uller and Woltzenlogel-Paleo \shortcite{C40} with the LEO-II prover. 
  The experiments reported here led to the discovery of a surprisingly accessible
  understanding of this inconsistency which is philosophically
  profound and which has not been presented yet in the
  literature. The detection of this inconsistency in combination with
  the work reported here thus demonstrates that artificial intelligence systems -- 
  particularly higher-order automated
  theorem provers -- are capable of assisting in the discovery and ellucidation of
  \emph{new} and philosophically relevant knowledge. 
\item Another open question has been whether the final theorem T3 \textit{``Necessarily, there
    exists God''} could eventually be proved fully automatically
  directly from the axioms alone. In \shortcite{C40}, the provers had only succeeded in proving 
  T3 from intermediate argumentation steps (lemmata). Here we report that, by using a more efficient new
  embedding for the higher-order modal logic S5, T3 can be directly proven from Scott's version of the axioms, which are consistent. The fully automatic proof has been
  generated by the theorem prover \textsc{Leo-II}~\cite{C26} and
  subsequently verified in the proof assistant
  Isabelle/HOL~\cite{NPW02}.
\item A third open challenge the verification in Isabelle/HOL of the modal
  collapse \cite{Sobel}, which is one of the most strongly criticised
  `side-effects' of G\"odel's and Scott's variants of the proof. In previous work
  the modal collapse has been derived by the provers
  \textsc{Satallax} \cite{Satallax} and \textsc{Leo-II}, but a fully automatic
  verification in the highly trusted Isabelle/HOL still failed. 
  Now, with the more efficient embedding for S5, this verification has been done. 
\item On the technical side, we present a significantly improved
  syntax representation of G\"odel's resp.~Scott's proof in
  Isabelle/HOL. In fact, we demonstrate that a nearly perfect match
  between the original pen and paper presentations and our encoding in
  Isabelle/HOl is feasible. This is clearly an important prerequisite
  for promoting the theorem proving technology employed here to a wide community of
  philosophers, who are not necessarily experts in automated reasoning or higher-order logic.
\item On the other hand, we
  also point to several technical challenges which, at least for the
  moment, still restrain non-expert users of these systems from
  independently conducting similar experiments in computational
  metaphysics as reported here and in previous work. 
% Technical issues detected within Isabelle-Sledgehammer; room for improvements
\end{itemize}

@@

@@

@@

>>>>

\begin{abstract}
  This paper discusses the discovery of the inconsistency in G\"odel's ontological argument as a success story for artificial intelligence. Despite the popularity of the argument since the appearance of G\"odel's manuscript in 1970, the inconsistency of the axioms used in the argument remained unnoticed until 2013, when it was detected automatically by the higher-order theorem prover LEO-II. Understanding and verifying the refutation generated by the prover turned out to be a time-consuming task, and its completion, as reported here, required the reconstruction of the refutation in the Isabelle proof assistant. The development of an improved syntactical hiding for the embedding technique, utilizing Isabelle's binding notation mechanism, allows the refutation to be presented in a human-friendly way, suitable for non-experts in the technicalities of higher-order theorem proving. This brings us a step closer to wider adoption of logic-based artificial intelligence tools by philosophers.
\end{abstract}

@@

The main novel contribution reported in this paper is a detailed analysis (in various modal logics) 
of the inconsistency of G\"{o}del's original version of the axioms used in his proof 
\shortcite{GoedelNotes}. The extraction, reconstruction and verification of an informal, 
human intuitive argument has been an open problem since the first detection of this inconsistency 
by Benzm\"uller and Woltzenlogel-Paleo \shortcite{C40} with the LEO-II prover. 
The verified refutation (discussed in Section \ref{ToDo}) displays a surprisingly accessible
explanation of the inconsistency, which is philosophically
profound and never presented in the
literature. The detection of this inconsistency in combination with
the work reported here thus demonstrates that artificial intelligence systems 
-- particularly higher-order automated theorem provers -- 
are capable of assisting in the discovery and elucidation of
\emph{new} and philosophically relevant knowledge.

On the technical side, the quest for constructing a compelling refutation, 
capable of convincing also human non-experts, led us to improve the 
syntax of the embedding of modal logics in
Isabelle/HOL (as discussed in Section  \ref{ToDo}). With the new syntax, a nearly perfect match
between the original pen and paper presentations and our encoding in
Isabelle/HOl is feasible. A more user-friendly syntax, as reported here, is clearly an important prerequisite
for promoting the theorem proving technology employed here to a wider community of
philosophers, who are not necessarily experts in automated reasoning or higher-order logic.

Another novel contribution reported here (in Section \ref{ToDo}) is the implementation of an alternative embedding for the only slightly stronger modal logic \SFiveU. Our experiments have shown that the new embedding is more efficient, as the following two open problems can now be solved:
\begin{itemize}
\item Proving the final theorem T3 \textit{``Necessarily, there
    exists God''} fully automatically and directly from the axioms alone, without relying on the argument's lemmas.
\item Verifying automatically the proof of the modal
  collapse \cite{Sobel}, which is one of the most strongly criticized
  `side-effects' of the argument's axioms.
\end{itemize}

\subsection{Related Work}

@@

Another open question has been whether the final theorem T3 \textit{``Necessarily, there
    exists God''} could eventually be proved fully automatically
  directly from the axioms alone. In \shortcite{C40}, the provers had only succeeded in proving 
  T3 from intermediate argumentation steps (lemmata). Here we report that, by using a more efficient new
  embedding for the higher-order modal logic S5, T3 can be directly proven from Scott's version of the axioms, which are consistent. The fully automatic proof has been
  generated by the theorem prover \textsc{Leo-II}~\cite{C26} and
  subsequently verified in the proof assistant
  Isabelle/HOL~\cite{NPW02}.

A third open challenge the verification in Isabelle/HOL of the modal
  collapse \cite{Sobel}, which is one of the most strongly criticised
  `side-effects' of G\"odel's and Scott's variants of the proof. In previous work
  the modal collapse has been derived by the provers
  \textsc{Satallax} \cite{Satallax} and \textsc{Leo-II}, but a fully automatic
  verification in the highly trusted Isabelle/HOL still failed. 
  Now, with the more efficient embedding for S5, this verification has been done.

\subsection{Improved Syntax in Isabelle}

use of binder notation, priorities

@@

is presented
  for higher-order modal logics KB and K (with possibilist and
  actualist first-order quantifiers)

@@

We also point to several technical challenges which, at least for the
moment, still restrain non-expert users of these systems from
independently conducting similar experiments in computational
metaphysics as reported here and in previous work.
% Technical issues detected within Isabelle-Sledgehammer; room for improvements
